A Domain-Specific Optimizing Compiler\\for Solving Partial Differential Equations

### 2014-02-25 paper/bare_conf.tex
Minor changes

% Autovectorizable LICM -> show code snippet?
\subsection{Auto-vectorizable Loop-invariant Code Motion}
From inspection of the codes in Figures~\ref{code:helmholtz} and~\ref{code:burgers}, it can be noticed that the computation of $A$ involves evaluating many sub-expressions that depend on two iteration variables only. Since symbols in most of these sub-expressions are read-only variables, there is ample space for loop-invariant code motion ($licm$). Vendor compilers apply $licm$, although not in the systematic way we need in our assembly kernels. We want to overcome two deficiencies that both $icc$ and $gcc$ have. First, these compilers can identify sub-expressions that are invariant with respect to the innermost loop only. This is an issue for sub-expressions depending on $i$-$k$ only, which are not automatically lifted. Second, the hoisted code is scalar, i.e. it is not subjected to auto-vectorization. We work around these limitations by doing source-level $licm$. In addition, we pre-compute all values that an invariant sub-expression assumes along the fastest varying dimension. This is implemented by introducing temporary arrays and new loops in the nest. At the price of both extra memory for storing temporaries and reduced register locality, the gain is that lifted terms can be auto-vectorized, because part of an inner loop. Note that given the short trip counts of our loops, it is important to achieve auto-vectorization of hoisted terms in order to minimize the percentage of scalar instructions, which can be otherwise significant.

% Padding and Data Alignment
\subsection{Padding and Data Alignment}
Auto-vectorization of assembly code that computes $A$ can be less effective if data are not aligned and if the length of the innermost loop is smaller than the vector length ($vl$). Data alignment is enforced in two steps. Initially, both arrays and matrices are allocated to addresses that are multiples of $vl$. Then, matrices are padded by rounding the number of columns to the nearest multiple of $vl$. For example, assume the original size of a matrix is 3$\times$3, and the underlying machine possesses AVX, which implies $vl=4$ since a vector register is 256 bits long and our kernels use double-precision floating-point values (64 bits). Then, a padded matrix on this architecture will be 3$\times$4. The compiler is explicitly informed about data alignment using a suitable pragma. Padding of all matrices involved in the evaluation of the stiffness matrix allows us to safely round the loop trip count to the nearest multiple of $vl$. This avoids the introduction of a remainder (scalar) loop from the compiler, which would be responsible for inefficient vectorization.

% Register allocation and register locality
\subsection{Toward Improving Register Locality}
One notable problem assembly kernels are exposed to concerns register allocation and register locality. The critical situation occurs when loop trip counts and number of accessed matrices along the inner dimensions are such that available vector registers are under-utilized. Due to a kernel's working set fitting the L1 cache, it is remarkably important to optimize register management in order to maximize performance. Canonical optimizations, such as loop interchange, unroll, unroll-and-jam and register tiling are typically employed to deal with this problem. In our compiler we support these optimizations either by means of explicit source code transformation (interchange, register tiling, unroll-and-jam) or indirectly advising the compiler through standard pragmas (unroll). With register tiling, in particular, we slice the innermost loop into rectangular blocks of identical size (except the ``reminder'' block), and we apply unrolling to individual blocks to expose ILP.

% remember: op vectorization PLUS unroll and jam !!
\subsection{Outer-product Vectorization}
Register tiles along both inner dimensions can be determined as well. Based on the observation that the evaluation of the stiffness matrix can be reduced to a summation of outer products along the $j$ and $k$ dimensions, aims at optimizing register reuse, in terms of both minimization of load operations and achievement of larger tiles, over a canonical implementation. If we consider the code snippet in Figure~\ref{code:helmholtz-licm}, we can notice that the computation of $A$ is abstractly expressible as
\begin{equation}
\label{outer-product}
A_{jk} = \sum_{\substack{
  x \in B' \subseteq B \\ 
  y \in B'' \subseteq B}} 
x_j\cdot y_k ~~~~~~ j,k = 0,...,4
\end{equation}
where $B$ is the set of all matrices (or temporaries) accessed in the kernel, whereas $B'$ and $B''$ are generic problem-dependent subsets. Without loss of generality, the presence of other constant or $i$-dependent variables can be momentarily neglected. Note that regardless of the specific input problem, the assembly computation is always reducible to this form. Figure~\ref{fig:vect-by-vect} illustrates how we can evaluate 16 elements ($j,k=0,...,4$) of the stiffness matrix with just 2 vector registers. Once the first four entries are evaluated, values in a register are shuffled and the same product is performed again. Standard compiler auto-vectorization ($gcc$ and $icc$), instead, would employ 5 registers, 4 of them target of a broadcast operation (i.e. ``splatting'' a single value over the four locations of an AVX register) along the outer dimension, and 1 of a load operation along the inner dimension. In our outer-product vectorization ($op$-$vect$), the 4 broadcast accesses to the L1 cache are avoided by an initial load and by shuffling. More importantly, we avoid using $vl$-1 registers per each outer-loop-dependent variable (i.e. all $x$ terms in Equation~\ref{outer-product}), with $vl$ being the vector length. This is a considerable gain, which allows us to slice the iteration space into bigger tiles, implemented directly by vector registers.

\begin{figure}[h]
\centerline{\includegraphics[scale=0.6]{Pictures/vect-by-vect.pdf}
\label{fig:vect-by-vect}}
\caption{Outer-product vectorization by permuting values in a vector register.}
\end{figure}

The storage layout of $A$, however, is incorrect after the application of $op$-$vect$. We efficiently restore it with a sequence of vector shuffle following the pattern highlighted in Figure~\ref{fig:restore-layout}, executed just once outside of the $ijk$ loop nest. The generated pseudo-code for the simple Helmholtz problem when using $op$-$vect$ is shown in Figure~\ref{code:helmholtz-opvect}.

\begin{figure}[h]
\centerline{\includegraphics[scale=0.7]{Pictures/vect-restore.pdf}
\label{fig:restore-layout}}
\caption{Restoring the storage layout after $op$-$vect$.}
\end{figure}

@@

% Splitting
\subsection{Assembly splitting}
In complex kernels, like Burgers in Figure~\ref{code:burgers}, and on certain architectures, achieving effective register allocation can be challenging. If the number of variables independent of the innermost-loop dimension, i.e. basis function matrices, temporaries introduced by $licm$ and constants, is larger than the available registers on the CPU, it is likely to obtain poor register reuse. For example, applying $licm$ to Burgers on a 3D mesh needs 33 temporaries for the $ijk$ loop order, and compiler's hoisting of invariant loads out of the $k$ loop can be inefficient on architectures with a relatively low number of registers. One potential solution to this problem consists of suitably ``splitting'' the computation of the stiffness matrix $A$ into multiple sub-expressions; an example, for the simpler Helmholtz problem, is given in Figure~\ref{code:helmholtz-split}. Splitting an expression has, in general, several drawbacks. Firstly, it increases the number of accesses to $A$ proportially to the ``split factor'', which is the number of sub-expressions produced. Secondly, depending on how the split is executed, it can lead to redundant computation (e.g. the product $det*W3[i]$ is performed times number of sub-expressions in the code of Figure~\ref{code:helmholtz-split}). Finally, it might affect register locality, although this is not the case of the Helmholtz example: for instance, the same matrix could be accessed in different sub-expressions, requiring a proportional number of loads be performed. Nevertheless, as shown in Section~\ref{sec:perf-results}, the performance gain from improved register reuse along inner dimensions can still be greater, especially if the split factor and the splitting itself use heuristics to minimize the aforementioned issues.

\begin{algorithm}[t]
\small
\caption{Local assembly code generated by Firedrake for a Helmholtz problem (2D mesh, Lagrange $p=2$ elements). The padding, data alignment, $licm$ and $op$-$vect$ optimizations are applied. Data alignment and padding relate to an AVX machine. In this specific case, the unroll-and-jam factor is 1.}
\label{code:helmholtz-opvect}
\KwSty{void} helmholtz(double A[8][8], double **coords) $\lbrace$\\
~~// K, det = Compute Jacobian (coords) \\
~~// Declaration of basis function matrices \\
~~\\
~~\KwSty{for} (\KwSty{int} i = 0; i$<$6; i++) $\lbrace$ \\
~~~~// Do loop-invariant code motion \\
~~~~\KwSty{for} (\KwSty{int} j = 0; j$<$4; j+=4) \\
~~~~~~\KwSty{for} (\KwSty{int} k = 0; k$<$8; k+=4) $\lbrace$\\
~~~~~~~~// Call $Load$ and $set$ intrinsics \\
~~~~~~~~// Compute A[1,1],A[2,2],A[3,3],A[4,4] \\
~~~~~~~~// One $permute\_pd$ intrinsics per $k$-loop $load$\\
~~~~~~~~// Compute A[1,2],A[2,1],A[3,4],A[4,3] \\
~~~~~~~~// One $permute2f128\_pd$ intrinsics per $k$-loop $load$\\
~~~~~~~~// ...\\
~~~~~~$\rbrace$\\
~~~~// Do Remainder loop (from $j=4$ to $j=6$)\\
~~$\rbrace$\\
~~// Restore the storage layout:\\
~~\KwSty{for} (\KwSty{int} j = 0; j$<$4; j+=4) $\lbrace$\\
~~~~$\_\_$m256d r0, r1, r2, r3, r4, r5, r6, r7;\\
~~~~\KwSty{for} (\KwSty{int} k = 0; k$<$8; k+=4) $\lbrace$\\
~~~~~~r0 = $\_$mm256$\_$load$\_$pd ($\&$A[j+0][k]);\\
~~~~~~// Load A[j+1][k], A[j+2][k], A[j+3][k]\\
~~~~~~r4 = $\_$mm256$\_$unpackhi$\_$pd (r1, r0);\\
~~~~~~r5 = $\_$mm256$\_$unpacklo$\_$pd (r0, r1);\\
~~~~~~r6 = $\_$mm256$\_$unpackhi$\_$pd (r2, r3);\\
~~~~~~r7 = $\_$mm256$\_$unpacklo$\_$pd (r3, r2);\\
~~~~~~r0 = $\_$mm256$\_$permute2f128$\_$pd (r5, r7, 32);\\
~~~~~~r1 = $\_$mm256$\_$permute2f128$\_$pd (r4, r6, 32);\\
~~~~~~r2 = $\_$mm256$\_$permute2f128$\_$pd (r7, r5, 49);\\
~~~~~~r3 = $\_$mm256$\_$permute2f128$\_$pd (r6, r4, 49);\\
~~~~~~$\_$mm256$\_$store$\_$pd ($\&$A[j+0][k], r0);\\
~~~~~~// Store A[j+1][k], A[j+2][k], A[j+3][k]\\
~~~~$\rbrace$\\
~~$\rbrace$\\
$\rbrace$
\end{algorithm}

@@

Firedrake provides users with the Unified Form Language to write problems in a notation resembling mathematical equations. This high-level specification is translated by the domain-specific Fenics Form Compiler~\cite{ffc} into an Abstract Syntax Tree representation of a Finite Element assembly kernel. ASTs are then passed to PyOP2, where parallel execution over the unstructured mesh is managed. Our compiler, capable of applying the transformations described in Section~\ref{sec:code-transf}, is integrated with PyOP2: it can manipulate ASTs introducing optimizations, and then it generates C code, which is eventually just-in-time compiled on the underlying architecture.

@@

Our experience... unrolling/unroll and jam

>>>>

% Padding and Data Alignment
\subsection{Padding and Data Alignment}
Auto-vectorization of assembly code computing the stiffness matrix can be less effective if data are not aligned and if the length of the innermost loop is smaller than the vector length ($vl$). Data alignment is enforced in two steps. Initially, both arrays and matrices are allocated to addresses that are multiples of $vl$. Then, matrices are padded by rounding the number of columns to the nearest multiple of $vl$. For example, assume the original size of a matrix is 3$\times$3 and that the underlying machine possesses AVX, which implies $vl=4$ since a vector register is 256 bits long and our kernels use 64-bits double-precision floating-point values. Then, a padded matrix on this architecture will have size 3$\times$4. The compiler is explicitly informed about data alignment using a suitable pragma. Padding of all matrices involved in the evaluation of the stiffness matrix allows us to safely round the loop trip count to the nearest multiple of $vl$. This avoids the introduction of a remainder (scalar) loop from the compiler, which would be responsible for inefficient vectorization.

% Autovectorizable LICM -> show code snippet?
\subsection{Auto-vectorizable Loop-invariant Code Motion}
From inspection of the codes in Figures~\ref{code:helmholtz} and~\ref{code:burgers}, it can be noticed that the computation of $A$ involves evaluating many sub-expressions that depend on two iteration variables only. Since symbols in most of these sub-expressions are read-only variables, there is ample space for loop-invariant code motion. Vendor compilers apply this technique, although not in the systematic way we need for our assembly kernels. We want to overcome two deficiencies that both $icc$ and $gcc$ have. First, these compilers can identify sub-expressions that are invariant with respect to the innermost loop only. This is an issue for sub-expressions depending on $i$-$k$, which are not automatically lifted. Second, the hoisted code is scalar, i.e. it is not subjected to auto-vectorization. We work around these limitations with source-level loop-invariant code motion. In particular, we pre-compute all values that an invariant sub-expression assumes along the fastest varying dimension. This is implemented by introducing a temporary array (per invariant sub-expression) and by adding a new loop to the nest. At the price of extra memory for storing temporaries, the gain is that lifted terms can be auto-vectorized, because part of an inner loop. Note that given the short trip counts of our loops, it is important to achieve auto-vectorization of all hoisted terms in order to minimize the percentage of scalar instructions, which could be otherwise significant.

Figure~\ref{code:helmholtz-licm} shows the Helmholtz assembly code after the application of loop-invariant code motion, padding, and data alignment.

% Register allocation and register locality
\subsection{Optimizing Register Management}
One notable problem assembly kernels are exposed to concerns register allocation and register locality. The critical situation occurs when loop trip counts and accessed variables are such that the vector-registers pressure is high. Due to a kernel's working set fitting the L1 cache, it is remarkably important to optimize register management in order to maximize performance. Canonical optimizations, such as loop interchange, unroll, unroll-and-jam and register tiling are typically employed to deal with this problem. In the compiler we have developed, we support these optimizations either by means of explicit source code transformation (interchange, unroll-and-jam) or indirectly advising the compiler through standard pragmas (unroll). 
In this context, register tiling is a critical optimization. A way of obtaining register tiles consists of slicing the innermost loop into rectangular blocks of identical size (except the ``reminder'' block), and applying unrolling to individual blocks in order to expose ILP. This is automated by some vendor compilers~\cite{comdex}. Register tiles along both inner dimensions can be determined as well. Based on the observation that the evaluation of the stiffness matrix can be reduced to a ``summation of outer products'' along the $j$ and $k$ dimensions, we implement a domain-specific vector-register tiling strategy that leads to larger tiles, enhancing reuse. If we consider the code snippet in Figure~\ref{code:helmholtz-licm} (Helmholtz after loop-invariant code motion), we can notice that the computation of $A$ is abstractly expressible as
\begin{equation}
\label{outer-product}
A_{jk} = \sum_{\substack{
  x \in B' \subseteq B \\ 
  y \in B'' \subseteq B}} 
x_j\cdot y_k ~~~~~~ j,k = 0,...,4
\end{equation}
where $B$ is the set of all matrices (or temporaries) accessed in the kernel, whereas $B'$ and $B''$ are generic problem-dependent subsets. Without loss of generality, the presence of constants or other variables independent of both $j$ and $k$ can be momentarily neglected. Note that regardless of the specific input problem, the stiffness matrix computation is always reducible to this kind of form. Figure~\ref{fig:vect-by-vect} illustrates how we can evaluate 16 elements ($j,k=0,...,4$) of the stiffness matrix using just 2 vector registers (a 4$\times$4 tile), assuming $\vert B' \vert = \vert B'' \vert = 1$. Values in a register are shuffled each time a product is performed. Standard compiler auto-vectorization ($gcc$ and $icc$), instead, executes 4 broadcast operations (i.e. ``splat'' of a value over all of the register locations) along the outer dimension to perform the calculation, and would also need to keep between $f=1$ and $f=3$ extra registers to perform the same 16 evaluations if unroll-and-jam is found to be useful, with $f$ being the unroll-and-jam factor.

%More importantly, we avoid using $vl$-1 registers per each outer-loop-dependent variable (i.e. all $x$ terms in Equation~\ref{outer-product}), with $vl$ being the vector length. This is a considerable gain, which allows us to slice the iteration space into bigger tiles, implemented directly by vector registers.

\begin{figure}[h]
\normalsize
\centerline{\includegraphics[scale=0.6]{Pictures/vect-by-vect.pdf}
\label{fig:vect-by-vect}}
\caption{Outer-product vectorization by permuting values in a vector register.}
\end{figure}

The storage layout of $A$, however, is incorrect after the application of this ``outer-product vectorization'' ($op$-$vect$). We efficiently restore it with a sequence of vector shuffle following the pattern highlighted in Figure~\ref{fig:restore-layout}, executed once outside of the $ijk$ loop nest. The generated pseudo-code for the simple Helmholtz problem when using $op$-$vect$ is shown in Figure~\ref{code:helmholtz-opvect}.

\begin{figure}[t]
\centerline{\includegraphics[scale=0.7]{Pictures/vect-restore.pdf}
\label{fig:restore-layout}}
\caption{Restoring the storage layout after $op$-$vect$. The figure shows how 4$\times$4 elements in the top-left block of the stiffness matrix $A$ can be brought to their correct positions.}
\end{figure}

@@

% Splitting
\subsection{Assembly splitting}
In complex kernels, like Burgers in Figure~\ref{code:burgers}, and on certain architectures, achieving effective register allocation can be challenging. If the number of variables independent of the innermost-loop dimension, i.e. basis function matrices, temporaries introduced by loop-invariant code motion and constants, is even close to the number of available CPU registers, it is likely to obtain poor register reuse. For example, applying loop-invariant code motion to Burgers on a 3D mesh needs 33 temporaries for the $ijk$ loop order, and compiler's hoisting of invariant loads out of the $k$ loop can be inefficient on architectures with a relatively low number of registers. One potential solution to this problem consists of suitably ``splitting'' the computation of the stiffness matrix $A$ into multiple sub-expressions; an example, for the simpler Helmholtz problem, is given in Figure~\ref{code:helmholtz-split}. Splitting an expression has, in general, several drawbacks. Firstly, it increases the number of accesses to $A$ proportially to the ``split factor'', which is the number of sub-expressions produced. Secondly, depending on how the split is executed, it can lead to redundant computation (e.g. the product $det*W3[i]$ is performed times number of sub-expressions in the code of Figure~\ref{code:helmholtz-split}). Finally, it might affect register locality, although this is not the case of the Helmholtz example: for instance, the same matrix could be accessed in different sub-expressions, requiring a proportional number of loads be performed. Nevertheless, as shown in Section~\ref{sec:perf-results}, the performance gain from improved register reuse along inner dimensions can still be greater, especially if the split factor and the splitting itself use heuristics to minimize the aforementioned issues.

\begin{algorithm}[t]
\small
\caption{Local assembly code generated by Firedrake for a Helmholtz problem (2D mesh, Lagrange $p=2$ elements). The padding, data alignment, $licm$ and $op$-$vect$ optimizations are applied. Data alignment and padding relate to an AVX machine. The original size of the $j$-$k$ iteration space (i.e. before padding was applied) was 6$\times$6. In this example, the unroll-and-jam factor is 1.}
\label{code:helmholtz-opvect}
\KwSty{void} helmholtz(double A[8][8], double **coords) $\lbrace$\\
~~// K, det = Compute Jacobian (coords) \\
~~// Declaration of basis function matrices \\
~~\\
~~\KwSty{for} (\KwSty{int} i = 0; i$<$6; i++) $\lbrace$ \\
~~~~// Do loop-invariant code motion \\
~~~~\KwSty{for} (\KwSty{int} j = 0; j$<$4; j+=4) \\
~~~~~~\KwSty{for} (\KwSty{int} k = 0; k$<$8; k+=4) $\lbrace$\\
~~~~~~~~// Call $Load$ and $set$ intrinsics \\
~~~~~~~~// Compute A[1,1],A[2,2],A[3,3],A[4,4] \\
~~~~~~~~// One $permute\_pd$ intrinsics per $k$-loop $load$\\
~~~~~~~~// Compute A[1,2],A[2,1],A[3,4],A[4,3] \\
~~~~~~~~// One $permute2f128\_pd$ intrinsics per $k$-loop $load$\\
~~~~~~~~// ...\\
~~~~~~$\rbrace$\\
~~~~// Do Remainder loop (from $j=4$ to $j=6$)\\
~~$\rbrace$\\
~~// Restore the storage layout:\\
~~\KwSty{for} (\KwSty{int} j = 0; j$<$4; j+=4) $\lbrace$\\
~~~~$\_\_$m256d r0, r1, r2, r3, r4, r5, r6, r7;\\
~~~~\KwSty{for} (\KwSty{int} k = 0; k$<$8; k+=4) $\lbrace$\\
~~~~~~r0 = $\_$mm256$\_$load$\_$pd ($\&$A[j+0][k]);\\
~~~~~~// Load A[j+1][k], A[j+2][k], A[j+3][k]\\
~~~~~~r4 = $\_$mm256$\_$unpackhi$\_$pd (r1, r0);\\
~~~~~~r5 = $\_$mm256$\_$unpacklo$\_$pd (r0, r1);\\
~~~~~~r6 = $\_$mm256$\_$unpackhi$\_$pd (r2, r3);\\
~~~~~~r7 = $\_$mm256$\_$unpacklo$\_$pd (r3, r2);\\
~~~~~~r0 = $\_$mm256$\_$permute2f128$\_$pd (r5, r7, 32);\\
~~~~~~r1 = $\_$mm256$\_$permute2f128$\_$pd (r4, r6, 32);\\
~~~~~~r2 = $\_$mm256$\_$permute2f128$\_$pd (r7, r5, 49);\\
~~~~~~r3 = $\_$mm256$\_$permute2f128$\_$pd (r6, r4, 49);\\
~~~~~~$\_$mm256$\_$store$\_$pd ($\&$A[j+0][k], r0);\\
~~~~~~// Store A[j+1][k], A[j+2][k], A[j+3][k]\\
~~~~$\rbrace$\\
~~$\rbrace$\\
$\rbrace$
\end{algorithm}

@@

Firedrake provides users with the Unified Form Language to write problems in a notation resembling mathematical equations. This high-level specification is translated by the domain-specific Fenics Form Compiler~\cite{ffc} into an Abstract Syntax Tree representation of a Finite Element assembly kernel. ASTs are then passed to PyOP2~\cite{pyop2isc}, which lies at the core of Firedrake, where parallel execution over the unstructured mesh is managed. Our compiler, capable of applying the transformations described in Section~\ref{sec:code-transf}, is integrated with PyOP2: it receives FFC's ASTs as input, it introduces optimizations, and it generates C code as output, which is eventually just-in-time compiled on the underlying architecture.

@@

Our experience... unrolling/unroll and jam/interchange
